The effects of host galaxy properties on merging compact binaries detectable by LIGO

### 2016-08-03 paper/mnras/paper.tex
last referee comments

\author[R. O'Shaughnessy et al.]{R. O'Shaughnessy,$^1$\thanks{E-mail: oshaughn@mail.rit.edu}
J. M. Bellovary,$^2$ 
A. Brooks$^3$
C. R. Christensen,$^4$ 
S. Shen,$^5$ 
F. Governato,$^6$ \\
%
%
%% List of institutions
$^1$Center for Computational Relativity and Gravitation, Rochester Institute of Technology, Rochester, NY 14623, USA  \\
$^2$Department of Physics, Queensborough Community College, Bayside, NY 11364, USA\\
$^3$Department of Physics and Astronomy, Rutgers University, Piscataway, New Jersey, 08854, USA\\
$^4$Department of Physics, Grinnell College, Grinnell, IA, 50112, USA\\
$^5$Institute of Astronomy, University of Cambridge, Cambridge CB3 OHA, UK\\
$^6$Department of Astronomy, University of Washington, Seattle, WA 98195, USA\\
 }

@@

This paper is organized as follows. 
In \S \ref{sec:sims}, we describe detailed hydrodynamical simulations of  several galaxies, including four of Milky Way-mass and two dwarfs.  Though the four Milky Way-mass galaxies are morphologically similar at $z = 0$, we
show that their star formation and chemical evolution history have subtle but important differences due to their
distinctive merger histories.   To demonstrate
these differences have a practical impact,  in \S \ref{sec:model}, we introduce a simple,
metallicity-dependent  phenomenological model to calculate the present-day rate and mass distribution of compact binary
mergers from  a galaxy's known history.   In \S \ref{sec:results:BBH}, we use this model to investigate  the compact
binary  coalescence rate
dependence on the galaxy's assembly history.   
%% We also demonstrate that even the coalescence rates of binary neutron stars (NSNS) and black hole-neutron
%% star (BH-NS) binaries also depend surprisingly sensitively on the assembly history of each galaxy. 
In \S~\ref{sec:Discussion} we discuss the implications of our study for the interpretation of host galaxy associations
identified via transient multmiessenger astronomy, in the short and long term.  We summarize our results in \S~ \ref{sec:conclude}.
%% In \S \ref{sec:results} we show that while these two galaxy formation models should produce comparable detectable binary neutron star
%% populations, these two scenarios could produce dramatically different populations of electromagnetically detectable
%% merging black hole-neutron star binaries. 
%

@@

We selected our simulated regions of interest from a volume of
uniform resolution, and resampled the region at very high
resolution using the volume renormalization technique \citep{Katz93}.
This technique allows us to follow the detailed physical processes
involved in galaxy evolution in our selected region while still
including large-scale torques from cosmic structure.  Our cosmological parameters are $\Omega_0 = 0.24$, $\Omega_{\rm baryon} = 0.04$, $\Lambda = 0.76$, h = 0.73, $\sigma_8 = 0.77$ 
 \citep{WMAP3} and we model the ionizing UV background
with the prescription from \citet{Haardt96}.  Our interstellar medium (ISM) model includes the non-equilibrium formation and destruction of H$_2$, which is incorporated in the cooling model along with metal lines \citep{Christensen12}.  Stars form
probabilistically from gas particles which meet density ($n_{min} =
0.1$ amu cm$^{-3}$) and temperature ($T_{max} = 10^3$ K) thresholds, though since star formation also depends on the H$_{\rm 2}$ content of a particle (see below) the densities are nearly always much higher than this threshold. If a gas
particle meets these criteria, it has a likelihood of forming a star
particle (representing a simple stellar population with a Kroupa IMF
\citep{Kroupa93}) which is given by

@@

@@

\begin{figure*}
\includegraphics[width=\columnwidth]{sfh}
\includegraphics[width=\columnwidth]{zvst}
\caption{\label{fig:TwoGalaxies}\textbf{Star formation and metallicity versus time}:  \emph{Left panel}: Star formation history $\dot{M}_{*}$
  versus time.  Black corresponds to \BoringGalaxy{} and red to \ExcitingGalaxy.  
\emph{Right panel}: A plot of the  metallicity $Z$ of recently-formed stars versus time.  Solid red and black lines show
the mean metallicity; dotted lines correspond to  90\% of newly-born stars with lower metallicity,   or 10\% of
newly-born stars with greater metallicity.  When these two galaxies are $2-3\unit{Gyr}$ old, prior to the first major
merger, newly-born stars are created with significantly different metallicity.  Additionally, prior to the
second major merger of h258 at $\simeq 6\unit{Gyr}$, stars form in the \ExcitingGalaxy{} galaxy at a systematically lower
metallicity than in the \BoringGalaxy{} counterpart.
}
\end{figure*}

@@

We have also employed the results of detailed simulations for two dwarf galaxies: \DwarfOne{} and \DwarfTwo{}.  The simulation of h603 consists of a low-mass disc galaxy (qualitatively similar to M33).  It has a virial mass of $3.4 \times 10^{11} \msun$, stellar mass of $7.8 \times 10^9 \msun$, and maximum circular velocity of 111 km s$^{-1}$.  The structure and star formation of this galaxy has been extensively studied by \citet{Christensen14b}.  We also include a bulgeless dwarf galaxy, h516, which has a disc with irregularly distributed star formation, with a virial mass of $3.8 \times 10^{10}\msun$, stellar mass $2.5 \times 10^8 \msun$, and maximum circular velocity of 65 km s$^{-1}$.  Images of these galaxies are shown in Figure \ref{fig:dwarfimages}, and we show their star formation history and metallicity evolution in  Figure \ref{fig:TwoDwarfGalaxies}.  Note that these galaxies have quite different masses, and are not meant to be directly comparable.  The more massive h603 has a much more active star formation history and an overall increasing metallicity with time, whereas the less massive h516 is characterized by small bursts of star formation and a fairly flat metallicity evolution.

@@

\begin{figure*}
\includegraphics[width=\columnwidth]{sfh_dwarf}
\includegraphics[width=\columnwidth]{dwarf_zvst}
\caption{\label{fig:TwoDwarfGalaxies}\textbf{Star formation and metallicity versus time}:  \emph{Left panel}: Star formation history $\dot{M}_{*}$
  versus time.  Red corresponds to h603 and black to h516.  
\emph{Right panel}: A plot of the  metallicity $Z$ of recently-formed stars versus time.  Solid red and black lines show
the mean metallicity; dotted lines correspond to  90\% of newly-born stars with lower metallicity,   or 10\% of
newly-born stars with greater metallicity.  %When these two galaxies are $2-3\unit{Gyr}$ old, prior to the first major
%merger of  near the peak of their
%star-forming history, newly-born stars are created with significantly different metallicity.  Additionally, prior to the
%second major merger of galaxy B at $\simeq 6\unit{Gyr}$, stars form in the \ExcitingGalaxy{} galaxy at a systematically lower
%metallicity than in the \BoringGalaxy{} counterpart.
}
\end{figure*}

@@

Our goal in this work is to estimate the \emph{ratio} of compact binaries that should be merging, at present, in the two
simulated Milky-Way analog galaxies described above as well as the dwarfs.

@@

%\editremark{recheck numbers: we are tweaking things, and Zsun is seemingly inconsistent in the two approaches }
 Our four Milky Way-like galaxies have extremely similar star formation histories and metallicity evolution, particularly at late
 times.  
However, at early times, the \ExcitingGalaxy{} galaxy forms stars for $\simeq 2\unit{Gyr}$ at a lower characteristic metallicity
($Z\simeq 0.8 \times10^{-3}$) compared
to the \BoringGalaxy{} galaxy ($Z\simeq 2\times 10^{-3}$).  In this regime, the formation efficiency $\lambda$ and sensitive volume $V$
can depend sensitively on mass; for example, for binary black holes, the ratio of  $(\lambda V)_{\BoringGalaxy{}}/(\lambda
V)_{\ExcitingGalaxy{}} \simeq 3$.\footnote{Adopting a more extreme exponent for the metallicity dependence ($a=2$) only
  changes this ratio by of order $2$.}   However, in this same epoch, the star formation rate in the \ExcitingGalaxy{} (low-metallicity)
galaxy is smaller, by a factor of roughly 2.
% 1e-3:  V = 475, lambda= 20   
% 2e-3:  V= 318, lambda = 10
% so the ratio should be a factor of about 7x higher due to this region.
Therefore, because only a fraction of order $10\%$  of all star formation occurs in this epoch, by this order of magnitude
argument, the overall number of
present-day coalescing binary black holes in our two galaxies is expected to differ by of order ten percent.
% 0.9 + 0.1*7/2
%
Using the concrete phenomenological calculations described above, we in fact find
$(r_{D}/M)_{\BoringGalaxy{}}/(r_{D}/M)_{\ExcitingGalaxy{}}\simeq 0.9$.    \jillianremark{\bf RICHARD this is 0.1 in my calcuations}

The close agreement between the two galaxies' binary black hole populations, determined by the anticorrelation between
star formation rate and metallicity, may be a single example of a broad trend.  This anticorrelation could cause
galaxies with similar present-day properties to always have similar present-day binary black hole populations,
regardless of their detailed assembly histories.
However, it is also possible that this close agreement is entirely a numerical coincidence between the specific star formation rate and metallicities in these two galaxies.  The effects described here
  are well within the scatter around the mass-metallicity relation.  Marginally different realizations of these
  histories can easily produce factors of order unity difference in galaxies with otherwise indistinguishable
  present-day properties.  That said, two other Milky-Way-like  galaxies (h239 and h285) also show very
  similar binary black hole coalescence rates $r_D/M$; see Table \ref{tab:Results}. % \editremark{but: do more examples show dramatic differences, or comparable changes}

More broadly, Table \ref{tab:Results} shows the results of our calculations for the three types of compact binaries
described above.   These calculations show just how dramatically different the compact binary populations of galaxies
with different present-day masses %but different star formation histories 
could be.   In particular, dwarf galaxies have an
exceptionally large fraction of low-metallicity star formation in their history \citep{Kirby13}.  The precise details of their chemical
evolution can modify their present-day binary black hole binary populations by factors of order a few, despite  adopting
the conservative choices described above for the dependence of rate on metallicity.

@@

For binary black hole mergers in particular, we show that the present-day binary black hole coalescence rate for our two
Milky-Way like galaxies is nearly identical, independent of their highly distinctive early-time formation histories.   
This result perhaps comes as somewhat of a surprise, considering the early differences in stellar and chemical evolution.
%% can vary by factors of several \jillianremark{ UPDATE} in otherwise
%% similar present-day environments. 
Our calculations adopt the same framework as  prior investigations, which demonstrated 
that black hole merger rates depend sensitively on low-metallicitiy environments
\citep[see,e.g.][]{PSellipticals,popsyn-LowMetallicityImpact2b-StarTrackRevised-2013,gwastro-EventPopsynPaper-2016}.   More broadly,
because compact binaries can merge long after they form, their host galaxy can evolve substantially in composition
between birth and merger.
%% and that the delay between compact
%% binary formation and coalesence ccn be
%% the sensitive dependence of binary evolution on
%% metallicity implies that
%% * BBH properties far from apparent.  No surprise: delays can be long , particularly because formation in low Z favored,
%% so system changes significantly from assembly conditions
Nonetheless, the apparent anticorrelation between the star formation rate and metallicity evolution of our galaxies has
led to similar late-time populations, despite substantial differences early on.
%
Further investigation is critical to assess whether this similarity is retained for more generic galaxy assembly
histories and binary formation models.    Now that we have introduced this method as a proof-of-concept here, it can be applied to large volume simulations, such as Illustris, EAGLE, or Romulus, in order to examine large numbers of galaxies and thus obtain statistically significant results.

>>>>

\author[R. O'Shaughnessy et al.]{R. O'Shaughnessy,$^1$\thanks{E-mail: oshaughn@mail.rit.edu}
J. M. Bellovary,$^2$ 
A. Brooks,$^3$
S. Shen,$^4$ 
F. Governato,$^5$ 
C. R. Christensen$^6$ \\
%
%
%% List of institutions
$^1$Center for Computational Relativity and Gravitation, Rochester Institute of Technology, Rochester, NY 14623, USA  \\
$^2$Department of Physics, Queensborough Community College, Bayside, NY 11364, USA\\
$^3$Department of Physics and Astronomy, Rutgers University, Piscataway, New Jersey, 08854, USA\\
$^4$Institute of Astronomy, University of Cambridge, Cambridge CB3 OHA, UK\\
$^5$Department of Astronomy, University of Washington, Seattle, WA 98195, USA\\
$^6$Department of Physics, Grinnell College, Grinnell, IA, 50112, USA\\
 }

@@

This paper is organized as follows. 
In \S \ref{sec:sims}, we describe detailed hydrodynamical simulations of  several galaxies, including four of Milky Way-mass and two dwarfs.  Though the four Milky Way-like galaxies are morphologically similar at $z = 0$, their star formation and chemical evolution history have subtle differences due to their
distinctive merger histories.   To demonstrate the practical impact of
these differences as well as that of halo mass,  in \S \ref{sec:model}, we introduce a simple,
metallicity-dependent  phenomenological model to calculate the present-day rate and mass distribution of compact binary
mergers from  a galaxy's known history.   In \S \ref{sec:results:BBH}, we use this model to investigate  the compact
binary  coalescence rate
dependence on each galaxy's assembly history and mass.   
%% We also demonstrate that even the coalescence rates of binary neutron stars (NSNS) and black hole-neutron
%% star (BH-NS) binaries also depend surprisingly sensitively on the assembly history of each galaxy. 
In \S~\ref{sec:Discussion} we discuss the implications of our study for the interpretation of host galaxy associations
identified via transient multmiessenger astronomy, in the short and long term.  We summarize our results in \S~ \ref{sec:conclude}.
%% In \S \ref{sec:results} we show that while these two galaxy formation models should produce comparable detectable binary neutron star
%% populations, these two scenarios could produce dramatically different populations of electromagnetically detectable
%% merging black hole-neutron star binaries. 
%

@@

We selected our simulated regions of interest from a volume of
uniform resolution, and resampled the region at very high
resolution using the volume renormalization technique \citep{Katz93}.
This technique allows us to follow the detailed physical processes
involved in galaxy evolution in our selected region while still
including large-scale torques from cosmic structure.  Our cosmological parameters are $\Omega_0 = 0.24$, $\Omega_{\rm baryon} = 0.04$, $\Lambda = 0.76$, h = 0.73, $\sigma_8 = 0.77$ 
 \citep{WMAP3} and we model the ionizing UV background
with the prescription from \citet{Haardt96}.  Our interstellar medium (ISM) model includes the non-equilibrium formation and destruction of H$_2$, which is incorporated in the cooling model along with metal lines, along with shielding of HI and H$_2$ \citep{Christensen12}.  Stars form
probabilistically from gas particles which meet density ($n_{min} =
0.1$ amu cm$^{-3}$) and temperature ($T_{max} = 10^3$ K) thresholds, though since star formation also depends on the H$_{\rm 2}$ content of a particle (see below) the densities are nearly always much higher than this threshold. If a gas
particle meets these criteria, it has a likelihood of forming a star
particle (representing a simple stellar population with a Kroupa IMF
\citep{Kroupa93}) which is given by

@@

We wish to point out that these simulations do not include the effects of supermassive black holes (SMBHs).  These galaxies are at the mass where feedback from SMBH accretion is thought to affect star formation, and adding these effects may add additional scatter to the stellar and metal evolution of each galaxy, which could alter the binary black hole merger history as well.

@@

\begin{figure*}
\includegraphics[width=\columnwidth]{sfh}
\includegraphics[width=\columnwidth]{zvst}
\caption{\label{fig:TwoGalaxies}\textbf{Star formation and metallicity versus time}:  \emph{Left panel}: Star formation history $\dot{M}_{*}$
  versus time.  Black corresponds to \BoringGalaxy{} and red to \ExcitingGalaxy.  
\emph{Right panel}: A plot of the  metallicity $Z$ of recently-formed stars versus time.  Solid red and black lines show
the mean metallicity; dotted lines correspond to  90\% of newly-born stars with lower metallicity,   or 10\% of
newly-born stars with greater metallicity.  The shaded region represents one standard deviation.  When these two galaxies are $2-3\unit{Gyr}$ old, prior to the first major
merger, newly-born stars are created with significantly different metallicity.  Additionally, prior to the
second major merger of h258 at $\simeq 6\unit{Gyr}$, stars form in the \ExcitingGalaxy{} galaxy at a systematically lower
metallicity than in the \BoringGalaxy{} counterpart.  Even from 8-13 Gyr, h258 has a slightly lower metallicity than h277.
}
\end{figure*}

@@

We have also employed the results of detailed simulations for two dwarf galaxies: \DwarfOne{} and \DwarfTwo{}.  The simulation of h603 consists of a low-mass disc galaxy (qualitatively similar to M33).  It has a virial mass of $3.4 \times 10^{11} \msun$, stellar mass of $7.8 \times 10^9 \msun$, and maximum circular velocity of 111 km s$^{-1}$.  The structure and star formation of this galaxy has been extensively studied by \citet{Christensen14b}.  We also include a bulgeless dwarf galaxy, h516, which has a disc with irregularly distributed star formation, with a virial mass of $3.8 \times 10^{10}\msun$, stellar mass $2.5 \times 10^8 \msun$, and maximum circular velocity of 65 km s$^{-1}$.  Images of these galaxies are shown in Figure \ref{fig:dwarfimages}, and we show their star formation history and metallicity evolution in  Figure \ref{fig:TwoDwarfGalaxies}.  Note that these galaxies have quite different masses, and are not meant to be directly comparable.  The more massive h603 has a much more active star formation history and an overall increasing metallicity with time, whereas the less massive h516 is characterized by small bursts of star formation and a fairly flat metallicity evolution, perhaps due to its substantial outflows.

@@

\begin{figure*}
\includegraphics[width=\columnwidth]{sfh_dwarf}
\includegraphics[width=\columnwidth]{dwarf_zvst}
\caption{\label{fig:TwoDwarfGalaxies}\textbf{Star formation and metallicity versus time}:  \emph{Left panel}: Star formation history $\dot{M}_{*}$
  versus time.  Red corresponds to h603 and black to h516.  
\emph{Right panel}: A plot of the  metallicity $Z$ of recently-formed stars versus time.  Solid red and black lines show
the mean metallicity; dotted lines correspond to  90\% of newly-born stars with lower metallicity,   or 10\% of
newly-born stars with greater metallicity.  The shaded region represents one standard deviation.%When these two galaxies are $2-3\unit{Gyr}$ old, prior to the first major
%merger of  near the peak of their
%star-forming history, newly-born stars are created with significantly different metallicity.  Additionally, prior to the
%second major merger of galaxy B at $\simeq 6\unit{Gyr}$, stars form in the \ExcitingGalaxy{} galaxy at a systematically lower
%metallicity than in the \BoringGalaxy{} counterpart.
}
\end{figure*}

@@

Our goal in this work is to estimate the \emph{ratio} of compact binaries that should be merging, at present, in the four
simulated Milky-Way analog galaxies described above as well as the dwarfs.

@@

%\editremark{recheck numbers: we are tweaking things, and Zsun is seemingly inconsistent in the two approaches }
 Our four Milky Way-like galaxies have extremely similar star formation histories and metallicity evolution, particularly at late
 times.  
However, at early times, the \ExcitingGalaxy{} galaxy forms stars for $\simeq 2\unit{Gyr}$ at a lower characteristic metallicity
($Z\simeq 0.8 \times10^{-3}$) compared
to the \BoringGalaxy{} galaxy ($Z\simeq 2\times 10^{-3}$).  In this regime, the formation efficiency $\lambda$ and sensitive volume $V$
can depend sensitively on mass; for example, for binary black holes, the ratio of  $(\lambda V)_{\BoringGalaxy{}}/(\lambda
V)_{\ExcitingGalaxy{}} \simeq 3$.\footnote{Adopting a more extreme exponent for the metallicity dependence ($a=2$) only
  changes this ratio by of order $2$.}   However, in this same epoch, the star formation rate in the \ExcitingGalaxy{} (low-metallicity)
galaxy is smaller, by a factor of roughly 2.
% 1e-3:  V = 475, lambda= 20   
% 2e-3:  V= 318, lambda = 10
% so the ratio should be a factor of about 7x higher due to this region.
Therefore, because only a fraction of order $10\%$  of all star formation occurs in this epoch, by this order of magnitude
argument, the overall number of
present-day coalescing binary black holes in our two galaxies is expected to differ by of order ten percent.
% 0.9 + 0.1*7/2
%
Using the concrete phenomenological calculations described above, we in fact find
$(r_{D}/M)_{\BoringGalaxy{}}/(r_{D}/M)_{\ExcitingGalaxy{}}\simeq 0.9$.

The close agreement between the two galaxies' binary black hole populations, determined by the anticorrelation between
star formation rate and metallicity, may be a single example of a broad trend.  This anticorrelation could cause
galaxies with similar present-day properties to always have similar present-day binary black hole populations,
regardless of their detailed assembly histories.  If low metallicity star formation makes up a small fraction of the total stellar mass, which is the case for each Milky Way-like galaxy we study here, then the quantity $V(Z) \lambda(Z) \dot{M}_* p(\log Z|t)$ must differ by about an order of magnitude for the binary black hole merger rates to differ strongly.  Such a scenario is possible if we alter our conservative choice of the exponent $a$ in the $\lambda(Z)$ function, but nonetheless is rather unlikely that galaxies in this mass range will undergo drastically different early evolution.
%However, it is also possible that this close agreement is entirely a numerical coincidence between the specific star formation rate and metallicities in these two galaxies.  
The effects described here
  are well within the scatter around the mass-metallicity relation.  Marginally different realizations of these
  histories can easily produce factors of order unity difference in galaxies with otherwise indistinguishable
  present-day properties.  %That said, two other Milky-Way-like  galaxies (h239 and h285) also show very
 % similar binary black hole coalescence rates $r_D/M$; see Table \ref{tab:Results}. % \editremark{but: do more examples show dramatic differences, or comparable changes}

More broadly, Table \ref{tab:Results} shows the results of our calculations for the three types of compact binaries
described above.   These calculations show just how dramatically {\em different} the compact binary populations of galaxies
with different present-day {\em masses} %but different star formation histories 
could be.   The dwarf galaxies have an
exceptionally large fraction of low-metallicity star formation in their history \citep{Kirby13} relative to the more massive galaxies.  The precise details of their chemical
evolution can modify their present-day binary black hole binary populations by factors of order a few, despite  adopting
the conservative choices described above for the dependence of rate on metallicity.  The progenitors and electromagnetic counterparts of future gravitational wave events are overall much more likely to be located in nearby dwarf galaxies with a large population of low metallicity stars.

@@

For binary black hole mergers in particular, we show that the present-day binary black hole coalescence rate for our two
Milky-Way like galaxies is nearly identical, independent of their highly distinctive early-time formation histories.   
This result perhaps comes as somewhat of a surprise, considering the early differences in stellar and chemical evolution.
%% can vary by factors of several \jillianremark{ UPDATE} in otherwise
%% similar present-day environments. 
Our calculations adopt the same framework as  prior investigations, which demonstrated 
that black hole merger rates depend sensitively on low-metallicitiy environments
\citep[see,e.g.][]{PSellipticals,popsyn-LowMetallicityImpact2b-StarTrackRevised-2013,gwastro-EventPopsynPaper-2016}.   More broadly,
because compact binaries can merge long after they form, their host galaxy can evolve substantially in composition
between birth and merger.
%% and that the delay between compact
%% binary formation and coalesence ccn be
%% the sensitive dependence of binary evolution on
%% metallicity implies that
%% * BBH properties far from apparent.  No surprise: delays can be long , particularly because formation in low Z favored,
%% so system changes significantly from assembly conditions
Nonetheless, the apparent anticorrelation between the star formation rate and metallicity evolution of our galaxies has
led to similar late-time populations, despite substantial differences early on.
%
Further investigation is critical to assess whether this similarity is retained for more generic galaxy assembly
histories and binary formation models.    Now that we have introduced this method as a proof-of-concept here, it can be applied to large volume simulations, such as Illustris, EAGLE, or Romulus \citep{Illustris,EAGLE,Tremmel16}, in order to examine large numbers of galaxies and thus obtain statistically significant results.
